---
title: "Environment Management in R with VerR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VerR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

In this vignette we will use VerR to compare two different version of an R package.
We will use the `dummyPkg` package, which is a dummy package that was created for this purpose.
The package has two versions, `0.1.1` and `0.1.2`, and we will compare them using VerR.
This package is available through GitHub, via `leopoldguyot/dummyPkg`.

# 1. Setup your environments

## Create a first environment
To start using VerR, we need to create a new environment.
To do so, we will use the `envCreate` function.
This function will create a new environment in the current working directory, 
and will populate it with the specified packages. Note that a new environment is 
just a new directory within `.envs/` that contains a `renv` project.

To specify the packages that we want to install in the environment, we have 
multiple options:

- **`pkg`**: Install the latest version of the package from CRAN.
- **`pkg@version`**: Install a specific version of the package from CRAN.
- **`username/repo`**: Install the package from GitHub.  
  You can also specify the commit with `username/repo@commitId`.
- **`bioc::pkg`**: Install the package from Bioconductor.

Note that to install a specific version of a Bioconductor package,
 it is recommended to install it via GitHub.

In our case we will use the GitHub option to install the `dummyPkg` package
 using his `sleep1` branch.
```{r envCreatePkg, message = FALSE}
library(VerR)

envCreate("sleep1Env", "leopoldguyot/dummyPkg@sleep1")
```

Once the environment is created, we can verify that it was correctly initialized.
To do so, we can use the `envList` function. It will list all the environments present
 in the `.envs/` directory.

```{r envList}
envList()
```

To obtain more details about this environment you can also use the `envInfo` 
function which will display extra information, like the installed packages in
the environment.

```{r envInfosleep1}
envInfo("sleep1Env", pkgInfo = TRUE)
```

## Extract the environment's lockFile

To save your environment configuration you can export his lockFile to a target
 location, this will allow to restore this environment later. 
 See `?renv::snapshot` for more information on lockFiles.
 To export this lockFile you can use the `lockFileExport` function.
 By default it will export the lockFiles of all available environments and
 store these files within the `exportedLockFiles/` directory inside your current
 working directory. You can also choose to select specific environments and 
 target paths for the lockFiles.

```{r lockFileExport}
lockFileExport()
```

The lockFile is now available in the `exportedLockFiles/` directory:
```{r checkLockFileExport}
list.files("exportedLockFiles")
```

## Create an environment from a lockFile

We will now use the exported lockFile to create a new environment. 
This environment will therefore be a clone of the `sleep1Env` environment.
To do so, we will again use the `envCreate` function, 
but this time with the `lockFile` argument.

```{r envCreateLockFile}
envCreate("sleep2Env", lockFile = "exportedLockFiles/sleep1Env_lockFile.lock")
```

Again we can check, that the environment was correctly created.

```{r envInfosleep2}
envInfo("sleep2Env", pkgInfo = TRUE)
```

## Update your environment

Currently we have two environments, `sleep1Env` and `sleep2Env`, that contain the 
same version of the `dummyPkg` package. We will now update the `sleep2Env` environment
to use another version of the `dummyPkg` package. To do so, we will use the
`envInstallPackage` function.

```{r envInstallPackage}
envInstallPackage(
    package = "leopoldguyot/dummyPkg@sleep2",
    envName = "sleep2Env"
)
```

To be sure that the package was correctly updated, we can check the installed packages
for both environments.

```{r envInfoBoth}
envInfo(pkgInfo = TRUE)
```

As expected, the `sleep2Env` environment now contains the `dummyPkg` package version `0.1.2` while
 the `sleep1Env` environment stills contain the version `0.1.1`.

# 2. Run expression within the environments

Now that we have two environments with different versions of the `dummyPkg` package,
 we can compare them using the `runInEnv` function. This function allows us to run
 an expression in a specific environment.

We will compare the `sleepy` function, which is a simple function that prints a message
 and sleeps for a certain amount of time. In the `sleep1Env` environment, the function
 sleeps for 1 second, while in the `sleep2Env` environment, it sleeps for 2 seconds.

```{r sleepy, eval = FALSE}
# version 0.1.1, x = 1
# version 0.1.2, x = 2
sleepy <- function() {
    message("Sleeping for x seconds")
    Sys.sleep(x)
}
```

```{r runInEnv}
runInEnv(
    expr = {
        library(dummyPkg)
        sleepy()
        return(packageVersion("dummyPkg"))
    },
    envName = envList()
)
```

The first thing to note is that any text output generated during the evaluation
 of the expression is displayed in the console. In this case, for the
 `sleep1Env` environment, the function prints `Sleeping for 1 seconds`, while
 for the `sleep2Env` environment, it prints `Sleeping for 2 seconds`.

The `runInEnv` function when called on multiple environments returns a named list,
 where the names correspond to the environment names, and the values are the return 
 values of the evaluated expression. As expected, the `sleep1Env` environment
 returns the version `0.1.1`, while the `sleep2Env` environment returns the
 version `0.1.2`.

# 4. Manipulate file through the environments

We will now explore how to send files to the environments,
 and also how to remove them.
For this example we will use an R script that contains code to
measure execution time of the sleepy function.
Here is the code contained in the script:

```{r scriptCode, eval = FALSE}
library("dummyPkg")

executionTimes <- replicate(3, {
    startTime <- Sys.time()
    sleepy()
    endTime <- Sys.time()
    as.numeric(difftime(endTime, startTime, units = "secs"))
})

executionTimes
```

We would want to run this script in both environments 
to compare the execution times.
First, we need to copy this script in each environment.
We use the `envCopyTo` function to copy the script to the environments.

```{r copyto}
scriptPath <- system.file("scripts",
    "sleepyBenchmark.R",
    package = "VerR"
)
envCopyTo(
    sourcePath = scriptPath,
    envName = envList(),
    targetPath = "scripts/sleepyBenchmark.R"
)
```

To verify that the copy was successful, we can use the
`fileInfo` parameter from the `envInfo ` function.
This will display a file tree of each environment:

```{r}
envInfo(fileInfo = TRUE)
```

# 5. Run a script in the environments

We will now run the script in both environments using the `runInEnv` function.
Using `source()` to run the script will return the last value of the script.
In this case, it will return the execution times of the `sleepy` function.

```{r}
result <- runInEnv(
    expr = {
        result <- source("scripts/sleepyBenchmark.R")
        result$value
    }
)
```

Lets now check the results:
```{r}
result

lapply(result, mean)
```

As expected, the average execution time for the `sleep1Env` environment is approximately 1 second, 
while for the `sleep2Env` environment, it is approximately 2 seconds.
This confirms that the sleepy function behaves as intended in both versions of the `dummyPkg` package.

# Benchmark in environments

`VerR` provides an easy way to benchmark the execution time of an expression in multiple
 environments.
The `benchInEnv` function takes an expression and measures the execution time of this expression
 in each environment. The `setup` parameter allow to evaluate R code before the 
 evaluation of the benchmarked expression, this is particularly useful to load
 packages or to retrieve data.
 You can also provide the `rep` argument to specify the number of replications
 to perform. The function will return a named list with the execution times for each environment.

```{r}
benchInEnv(
    expr = {
        sleepy()
    },
    envName = envList(),
    rep = 3,
    setup = {
        library(dummyPkg)
    },
    resultAggregation = mean
)
```

# Cleaning

```{r cleaning}
envDelete(force = TRUE)
unlink("exportedLockFiles",
    recursive = TRUE,
    force = TRUE
)
```
