---
title: "Environment Management in R with VerR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VerR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

In this vignette we will use VerR to compare two different version of an R package.
We will use the `dummyPkg` package, which is a dummy package that was created for this purpose.
The package has two versions, `0.1.1` and `0.1.2`, and we will compare them using VerR.
This package is available through GitHub, via `leopoldguyot/dummyPkg`.

# 1. Setup your environments

## Create a first environment
To start using VerR, we need to create a new environment.
To do so, we will use the `envCreate` function.
This function will create a new environment in the current working directory, 
and will populate it with the specified packages. Note that a new environment is 
just a new directory within `.envs/` that contains a `renv` project.

To specify the packages that we want to install in the environment, we have 
multiple options:

- **`pkg`**: Install the latest version of the package from CRAN.
- **`pkg@version`**: Install a specific version of the package from CRAN.
- **`username/repo`**: Install the package from GitHub.  
  You can also specify the commit with `username/repo@commitId`.
- **`bioc::pkg`**: Install the package from Bioconductor.

Note that to install a specific version of a Bioconductor package,
 it is recommended to install it via GitHub.

In our case we will use the GitHub option to install the `dummyPkg` package
 using his `sleep1` branch.
```{r envCreatePkg, message = FALSE, results=FALSE}
library(VerR)

envCreate("sleep1Env", "leopoldguyot/dummyPkg@sleep1")
```

Once the environment is created, we can verify that it was correctly initialized.
To do so, we can use the `envList` function. It will list all the environments present
 in the `.envs/` directory.

```{r envList}
envList()
```

To obtain more details about this environment you can also use the `envInfo` 
function which will display extra information, like the installed packages in
the environment.

```{r envInfosleep1}
envInfo("sleep1Env", pkgInfo = TRUE)
```

## Extracting the environment's lockFile

To save your environment configuration you can export his lockFile to a target
 location, this will allow to restore this environment later. 
 See `?renv::snapshot` for more information on lockFiles.
 To export this lockFile you can use the `lockFileExport` function.
 By default it will export the lockFiles of all available environments and
 store these files within the `exportedLockFiles/` directory inside your current
 working directory. You can also choose to select specific environments and 
 target paths for the lockFiles.

```{r lockFileExport}
lockFileExport()
```

The lockFile is now available in the `exportedLockFiles/` directory:
```{r checkLockFileExport}
list.files("exportedLockFiles")
```

## Create an environment from a lockFile

We will now use the exported lockFile to create a new environment. 
This environment will therefore be a clone of the `sleep1Env` environment.
To do so, we will again use the `envCreate` function, 
but this time with the `lockFile` argument.

```{r envCreateLockFile, results=FALSE}
envCreate("sleep2Env", lockFile = "exportedLockFiles/sleep1Env_lockFile.lock")
```

Again we can check, that the environment was correctly created.

```{r envInfosleep2}
envInfo("sleep2Env", pkgInfo = TRUE)
```

## Update your environment

Currently we have two environments, `sleep1Env` and `sleep2Env`, that contain the 
same version of the `dummyPkg` package. We will now update the `sleep2Env` environment
to use another version of the `dummyPkg` package. To do so, we will use the
`envInstallPackage` function.

```{r envInstallPackage}
envInstallPackage(
    package = "leopoldguyot/dummyPkg@sleep2",
    envName = "sleep2Env"
)
```

To be sure that the package was correctly updated, we can check the installed packages
for both environments.

```{r envInfoBoth}
envInfo(pkgInfo = TRUE)
```

As expected, the `sleep2Env` environment now contains the `dummyPkg` package version `0.1.2` while
 the `sleep1Env` environment stills contain the version `0.1.1`.
 
# Cleaning

```{r cleaning}
envDelete(force = TRUE)
unlink("exportedLockFiles",
    recursive = TRUE,
    force = TRUE
)
```

